---
title: "Cell Annotation -- Types & States"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    df_print: paged
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})
## supplied by annotate_data.R via the render environment:
##   ann_obj, ann_rules, ann_ct, ann_states, ann_plots_dir, ann_n_max
obj    <- ann_obj
rules  <- ann_rules
ct     <- ann_ct
states <- ann_states
pdir   <- ann_plots_dir
n_max  <- ann_n_max

## absolute, normalized path so include_graphics finds the PNG regardless of the
## knit working directory; self_contained output base64-embeds it. Returns NULL
## if the plot was not produced (e.g. no tumor cells), so chunks can skip it.
img <- function(name) {
  f <- file.path(pdir, name)
  if (!file.exists(f)) return(invisible(NULL))
  knitr::include_graphics(normalizePath(f, winslash = "/", mustWork = TRUE))
}
```

```{r intro-note, results="asis"}
cat("This report assigns each cell a **type** and **states** from the marker ",
    "positivity calls, using the rules shown below. ", sep = "")
if (identical(rules$conflict_resolution$policy, "priority")) {
  cat("Cells that match more than one cell type are settled by the priority ",
      "order set out in those rules; where that order picks no single winner ",
      "they are left as `UNKNOWN`.\n", sep = "")
} else {
  cat("It is a **first pass**: cells that match more than one cell type are ",
      "deliberately left as `UNKNOWN` (a conflict bucket) rather than resolved ",
      "by a priority order.\n", sep = "")
}
```

```{r rowcap-note, results="asis"}
if (is.finite(n_max)) {
  cat(sprintf(
    "> **Fast QC mode:** only the first **%s** rows of each Halo file were read. The composition and state counts below reflect only those sampled rows. Re-run with `--full` for whole-file numbers.\n",
    format(n_max, big.mark = ",")
  ))
}
```

# The rules these calls were made from

The same machine-readable rules file drives both the code and this description,
so what follows is exactly what was computed. The standalone copy is written
alongside this report as `cell_rules.md` for circulation.

```{r rules, results="asis"}
cat(HaloXi::rules_md_lines(rules), sep = "\n")
```

# Cell-type composition

Counts of each cell type per sample. `UNKNOWN` = positive for more than one
lineage's markers (conflict); `UNCLASSIFIED` = positive for no lineage marker;
`(NA / un-callable)` = a needed marker was absent from that sample.

```{r comp-tab}
ct$wide |> kable()
```

```{r comp-plot-counts, out.width="85%"}
img("composition_counts.png")
```

```{r comp-plot-pct, out.width="85%"}
img("composition_pct.png")
```

# Sub-states

For each state, the percentage of the relevant **parent-type** cells that carry
the flag (`nScored` is the parent-cell denominator). A blank percentage means
un-callable in that sample -- the marker was absent, or there were no parent
cells -- never a silent zero.

```{r states-tab}
states |> kable()
```

```{r states-plot, out.width="80%"}
img("state_heatmap.png")
```

## States by parent type

One panel per parent type the rules define states for: of that type's cells,
how many carry each flag (the flags are independent, not a partition).

```{r parent-state-files}
## same tag that names the state columns, so the panels come out in rules order
state_pngs <- file.path(
  pdir, sprintf("states_%s.png", HaloXi::state_tags(rules)))
state_pngs <- state_pngs[file.exists(state_pngs)]
```

```{r parent-state-note, results="asis"}
if (length(state_pngs) == 0) {
  cat("\n*No cells were called as any state parent type in this run, so there ",
      "are no per-type state results to show. In fast QC mode this is expected ",
      "when the sampled rows contain no cells matching those types.*\n", sep = "")
}
```

```{r parent-state-plots, out.width="75%"}
if (length(state_pngs) > 0) {
  knitr::include_graphics(normalizePath(state_pngs, winslash = "/"))
}
```

---

*Generated by HaloXi `annotate_data.R`. The composition and state tables are
also written as `Cell_annotation.xlsx`, and the rules as `cell_rules.md`,
alongside this report.*
