---
title: "Halo Data Scan -- Initial Overview"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    df_print: paged
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})
## `scan_res` (the scan result list) and `scan_plots_dir` (absolute path to the
## rendered PNGs) are supplied by scan_data.R via the render environment.
res <- scan_res
tab <- res$tables
meta <- res$meta
pdir <- scan_plots_dir
## absolute, normalized path so include_graphics finds the PNG regardless of
## the knit working directory; self_contained output base64-embeds it.
img <- function(name) {
  f <- normalizePath(file.path(pdir, name), winslash = "/", mustWork = TRUE)
  knitr::include_graphics(f)
}
```

This report is an **initial QC scan** of the Halo object files listed in the
sample manifest. It is meant to give a quick overview of data extent and
marker-panel coverage before any analysis -- not a finished result.

```{r rowcap-note, results="asis"}
if (is.finite(meta$n_max)) {
  cat(sprintf(
    "> **Fast QC mode:** only the first **%s** rows of each Halo file were read. Marker-panel coverage below is read from the full file headers and is complete, but cell counts and positivity percentages reflect only the sampled rows. Re-run with `--full` for exact whole-file numbers.\n",
    format(meta$n_max, big.mark = ",")
  ))
}
```

## Scan summary

```{r summary}
tibble::tibble(
  Field = c("Manifest", "Scanned at", "HaloXi version", "Rows read per file",
            "Samples in manifest", "Samples loaded",
            "Distinct markers", "Cells scanned"),
  Value = c(meta$manifest_csv, meta$scanned_at, meta$version,
            if (is.finite(meta$n_max)) format(meta$n_max, big.mark = ",") else "all rows",
            as.character(meta$n_samples), as.character(meta$n_loaded),
            as.character(meta$n_markers),
            format(meta$total_cells, big.mark = ","))
) |> kable()
```

## Manifest

```{r manifest}
tab$manifest |>
  mutate(HaloFile = basename(HaloFile)) |>
  kable()
```

## Cells per sample

"Phenotyped" cells are positive for at least one non-control marker (DAPI
excluded); the rest are negative for every marker in the panel. In fast QC mode
these counts are over the sampled rows, not the whole file.

```{r cellcounts-tab}
tab$cell_summary |>
  mutate(
    nCells = format(nCells, big.mark = ","),
    nPhenotyped = format(nPhenotyped, big.mark = ","),
    nUnphenotyped = format(nUnphenotyped, big.mark = ",")
  ) |>
  kable(align = "lrrrr")
```

```{r cellcounts-plot, out.width="80%"}
img("cell_counts.png")
```

## Marker panel coverage (sample x marker)

Which markers are present in each sample's panel. Panels are **not identical**
across samples -- markers absent from a sample are shown blank/grey. This
matters for any cross-sample comparison: a marker can only be compared where it
exists in every sample being compared.

```{r presence-tab}
tab$presence_matrix |> kable()
```

```{r presence-plot, out.width="70%"}
img("marker_presence.png")
```

## Marker positivity

Percentage of scored cells positive for each marker, per sample. Blank cells in
the matrix mean the marker is **absent from that sample's panel** (distinct from
present-but-zero). Control markers (e.g. DAPI) are flagged.

```{r positivity-tab}
tab$positivity_pct |> kable()
```

```{r positivity-plot, out.width="80%"}
img("marker_heatmap.png")
```

## Spatial footprint

Cell centroids per sample (down-sampled for plotting), coloured by whether the
cell is phenotyped. Useful for spotting tissue extent, empty regions, or
artefacts before spatial analysis.

```{r spatial-plot, out.width="100%"}
img("spatial_footprint.png")
```

---

*Generated by HaloXi `scan_data.R`. Excel workbooks with the full tables are
written alongside this report (`Halo_scan_overview.xlsx`,
`Halo_scan_markers.xlsx`).*
