---
title: "Spatial analysis -- co-location and neighbourhood"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    df_print: paged
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})
## supplied by spatial_data.R via the render environment:
##   sp_coloc, sp_neigh, sp_rules, sp_plots_dir, sp_n_max, sp_um_per_px,
##   sp_radii, sp_samples, sp_dup_centroids
coloc  <- sp_coloc
neigh  <- sp_neigh
pdir   <- sp_plots_dir
n_max  <- sp_n_max
upp    <- sp_um_per_px
radii  <- sp_radii
p      <- neigh$params
p_floor <- 2 / (p$n_perm + 1)

img <- function(name) {
  f <- file.path(pdir, name)
  if (!file.exists(f)) return(invisible(NULL))
  knitr::include_graphics(normalizePath(f, winslash = "/", mustWork = TRUE))
}

## a p at the floor is the resolution of the test, not a p of zero, so it is
## never printed as a number that reads like one
fmt_p <- function(x) {
  ifelse(x <= p_floor + 1e-12,
         paste0("<", format(signif(p_floor, 2))),
         format(signif(x, 3)))
}
```

```{r banner, results="asis"}
if (is.finite(n_max)) {
  cat(sprintf(
    "> **Fast QC mode:** only the first **%s** rows of each Halo file were read. Every count and percentage below reflects only those sampled rows. Re-run with `--full` for whole-file numbers.\n\n",
    format(n_max, big.mark = ",")
  ))
}
cat(sprintf(
  "> **Scan resolution: %s um per pixel.** Halo exports coordinates in pixels and carries no scan metadata, so every distance below is only as right as this number. It was supplied on the command line, not inferred.\n\n",
  format(upp)
))
cat(sprintf(
  "Specimens analysed: **%s**. Distances are measured centre to centre between cell bounding-box midpoints.\n",
  paste(sp_samples, collapse = ", ")
))
if (sp_dup_centroids > 0) {
  cat(sprintf(
    "\n%s cells share a centroid with another cell. Pairs at distance exactly 0 fall outside the half-open distance band and are not counted anywhere below.\n",
    format(sp_dup_centroids, big.mark = ",")
  ))
}
```

# How this is tested

Both analyses compare what is observed against **a random rearrangement of the
same cell-type labels over the same cells**, within each specimen, repeated
`r p$n_perm` times.

That null is the point. Tissue is not a uniform random field: cells are packed
into structures, and against a null that assumes uniformity nearly every pair
of types in a real section comes out significant. Shuffling the labels holds
the geometry, the local packing density and the specimen's overall composition
fixed, and destroys only the association between where a cell is and what it
is -- which is the thing being asked about.

Two consequences worth keeping in view:

- **The smallest p this can report is `r format(signif(p_floor, 2))`**
  (`2 / (permutations + 1)`). A row at that value means "smaller than this test
  resolves", never `p = 0`. Raising the permutation count is what resolves
  further.
- **Cells near the tissue edge have neighbourhoods cut off by the edge**, so
  raw counts there are biased low. The ratios -- `log2FC` and `log2Enrich` --
  are not affected, because the null is built over the same cells with the same
  truncation, so the bias cancels. No border correction is applied; read the
  ratios rather than the counts.

# Which cell types sit closer together than chance

Every pair of cell types, counted over cell pairs whose centres fall within
`r paste(radii, collapse = " and ")` um, against the shuffled-label null.
Self-pairs are included: a type paired with itself measures how clustered that
type is.

```{r coloc-plot}
img("colocation_heatmap.png")
```

## The strongest associations

```{r coloc-table}
coloc |>
  filter(qBH < 0.05) |>
  arrange(desc(abs(log2FC))) |>
  head(30) |>
  transmute(
    Band = paste0(RadiusInner, "-", RadiusOuter, " um"),
    Sample, TypeA, TypeB,
    nObs = round(nObs), nExp = round(nExp),
    log2FC = round(log2FC, 2), z = round(z, 1),
    pEmp = fmt_p(pEmp), qBH = signif(qBH, 3)
  ) |>
  kable(caption = "Type-pairs clearing q < 0.05, strongest effect first. nObs is the observed pair count, nExp what shuffling the labels gives.")
```

# What the tissue around each `r p$anchor_type` cell is made of

The ring is `r p$r_inner`-`r p$r_outer` um: a ring rather than a disc, so it
describes the surrounding tissue rather than the cells in immediate contact.
`r format(max(neigh$summary$nAnchors), big.mark = ",")` anchor cells at most,
per specimen.

```{r neigh-composition}
img("neighborhood_composition.png")
```

```{r neigh-enrichment}
img("neighborhood_enrichment.png")
```

## Ring composition

```{r neigh-table}
neigh$summary |>
  arrange(Sample, desc(pctOfRing)) |>
  transmute(
    Sample, NeighborType,
    nAnchors, nNeighbors,
    meanPerAnchor = round(meanPerAnchor, 2),
    pctOfRing = round(pctOfRing, 2),
    pctBackground = round(pctBackground, 2),
    log2Enrich = round(log2Enrich, 2),
    pEmp = fmt_p(pEmp), qBH = signif(qBH, 3)
  ) |>
  kable(caption = "pctOfRing is the type's share of the ring; pctBackground its share of the whole specimen. log2Enrich compares the ring against the shuffled-label null and is unaffected by rings cut off at the tissue edge.")
```

## How much the rings vary

`meanPerAnchor` is a mean over anchor cells whose rings differ a great deal --
some sit in dense tissue, some at an edge. The spread behind it:

```{r ring-spread}
neigh$per_anchor |>
  group_by(Sample) |>
  summarize(
    anchors = n(),
    mean = round(mean(nRing), 1),
    min = min(nRing),
    q25 = stats::quantile(nRing, 0.25),
    median = stats::median(nRing),
    q75 = stats::quantile(nRing, 0.75),
    max = max(nRing),
    .groups = "drop"
  ) |>
  kable(caption = "Cells in the ring, per anchor cell.")
```

# Parameters

```{r params}
tibble::tibble(
  Parameter = c("microns per pixel", "co-location radii (um)",
                "neighbourhood ring (um)", "anchor type",
                "permutations", "smallest reportable p", "seed",
                "row cap"),
  Value = c(
    format(upp),
    paste(radii, collapse = ", "),
    paste0(p$r_inner, "-", p$r_outer),
    p$anchor_type,
    format(p$n_perm),
    format(signif(p_floor, 2)),
    format(p$seed),
    if (is.finite(n_max)) format(n_max, big.mark = ",") else "all rows"
  )
) |> kable()
```
